Joint estimation of SNP effects improves the prediction of complex diseases // Predicting complex diseases: performance and robustness [OF NEW METHOD?]

### 2018-03-22 paper/draft.tex
Mike's changes

\newpage
[ABSTRACT TROP LONG?]
\abstract{
Polygenic Risk Scores (PRSs) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRSs can have a major public health impact, allowing for screening campaigns to identify high-genetic risk individuals for a given disease. Yet, current PRSs are far from reaching this goal, and would have to show a high discriminative power between cases and controls in order to be of any use.

The predictive power of PRS strongly relies on the underlying technique used to include SNPs in the PRS and to estimate their relative contribution. The most commonly used method to derive PRSs is called ``Clumping+Thresholding'' (C+T). It uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it very suitable and also very fast. We recently developed two R packages, bigstatsr and bigsnpr, for the analysis of large-scale genome-wide data. Package bigstatsr includes efficient functions for computing penalized linear and logistic regressions on large datasets.

We present a comprehensive comparative study of the C+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests and has shown exceptionally good predictive results in a previous study. We quickly discard the T-Trees method and show that the penalized logistic regression consistently achieves higher predictive performance while being as fast as the C+T method, whereas predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture, as shown in previous studies.\\
\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
\textbf{Supplementary information:}
}

@@

In order to simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}. We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only chromosome 6), the size of the training set ($\{1000, 2000, \dots, 6000\}$), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace). We also use two different models to compute liability scores: a ``simple'' model with linear effects only, and a ``fancy'' model which combines linear, dominant and interaction-type effects.
For the ``simple'' model, we compute $$y_i = \sum_{j\in E_\text{causal}} w_j \cdot \widetilde{G_{i,j}}$$ where $w_j$ are weights generated with a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$ and $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs). 
For the ``fancy'' model, we separate the causal SNPs in three equal sets $E_\text{causal}^{(1)}$, $E_\text{causal}^{(2)}$ 
and $E_\text{causal}^{(3)}$ ($E_\text{causal}^{(3)}$ is further separated in two equal sets, $E_\text{causal}^{(3.1)}$ and $E_\text{causal}^{(3.2)}$). 
We then compute $$y_i = \underbrace{\sum_{j\in E_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in E_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|E_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction}$$ 
where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $E_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|E_\text{causal}^{(q)}\right|\right\}\right\}$. 
Note that for the interaction part of the model, we scale interactions, not the raw allele counts, so that corresponding SNPs still have a marginal effect. 
For both models, the values of $y_i$ are then standardized so that they have a variance equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$, where $\epsilon \sim N(0, 1 - h^2)$.

@@

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of a Genome-Wide Associations Studies (GWAS), referred to as summary statistics.
For the C+T method, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part). 
The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another. 
Then, thresholding (T) consists in removing SNPs that are within a certain level of significance (p-value threshold to be determined). 
Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$S_i(T, E_\text{clumping}) = \sum_{j \in E_\text{clumping}} \mathds{1}\{p_j < T\} \cdot \beta_j \cdot G_{i,j}$$ where $\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS and $T$ is a threshold on p-values to be determined. In this study, we report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb.

For the C+T method, we report three different scores of prediction: one including all the SNPs (remaining after clumping), one including only SNPs that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$), and one that maximizes the AUC for these two thresholds and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}). 
Note that, normally, this optimal threshold would have to be learned based on the training set only in order to prevent from over-fitting the data. We consider this reported maximum AUC as an upper bound of the AUC for the C+T method.
We call these three reported scores respectively ``PRS-all'', ``PRS-stringent'' and ``PRS-max'' in the results.

T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests \cite[]{breiman2001random} that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium in GWAS data \cite[]{botta2014exploiting}. Since this method is ..significantly.. different from the other ones presented here and give that it achieved surprisingly high predictive performance in the past, we decided to include T-Trees in this comparison. We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}, except that we use 100 trees only (instead of 1000) because ..it.. showed only a minimal increase of AUC while requiring a disproportionately long processing time (e.g.\ AUC of 81.5\% instead of 81\%, data not shown). %We call this method ``T-Trees'' in the results.

The last method ..in/of.. this comparison is a penalized logistic regression.
We solve: $$\argmin_{\beta_0, \beta}(x, y, \lambda, \alpha)\left\{\underbrace{\frac{1}{n}\sum_{i=1}^n \log\left(1+e^{-y_i (\beta_0+x_i^T\beta)}\right)}_\text{Loss function} + \underbrace{\lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right)}_\text{Penalization}\right\}$$ where, in this study, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict and $\lambda$ is a regularization parameter that needs to be determined. 
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be exactly ..zero/equal to zero.. and can be used as a means of variable selection, leading to sparse (and therefore more interpretable) models  (corresponds to $\alpha = 1$); 
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. In this study, we always use the elastic-net regularization with $\alpha = 0.5$, without trying to tune (e.g.\ by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.

To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in the R package bigstatsr \cite[]{prive2017efficient}. 
This type of algorithm builds predictions for many values of $\lambda$ (typically a ``regularization path'' of 100 values). To obtain an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we called Cross-Model Selection and Averaging (CMSA). First, this procedure separates the training set in $K$ folds (e.g.\ 10 folds). 
Secondly, in turn, each fold is considered as an inner validation set and the other ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which maximizes the likelihood on the inner validation set is determined and the vector of coefficients corresponding to the previous vector of scores is selected. Finally, the $K$ resulting vectors of coefficients are reported and can be combined into one vector (e.g.\ by using the geometric median) or resulting scores can be combined (e.g.\ by averaging the $K$ probabilities for each individual). 
We also use this procedure to derive an early stopping criterion so that the algorithm does not need to evaluate the whole regularization paths, making this procedure much faster.
Because of the L1-regularization, the resulting vectors of coefficients are typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts. We refer to this method as ``logit-simple'' in the results section.

In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct a separate dataset with, for each SNP variable, two more variables coding for recessive and dominant effects. This results in a dataset with 3 times as many variables as the initial one, on which we can apply the penalized logistic regression with the CMSA procedure, as described previously. We call this method ``logit-triple'' in the results.

>>>>

\newpage
[ABSTRACT TROP LONG?]Yet, current PRSs are far from reaching this goal, and would have to show a high discriminative power between cases and controls in order to be of any use.
The predictive power of PRS strongly relies on the underlying technique used to include SNPs in the PRS and to estimate their relative contribution.

\abstract{
Polygenic Risk Scores (PRSs) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRSs can have a major public health impact, allowing for screening campaigns to identify high-genetic risk individuals for a given disease.  The most commonly used method to derive PRSs is called ``Clumping+Thresholding'' (C+T). It uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it very suitable and fast. However, PRS rely on non-optimal heuristics to choose various numerical thresholds.

We implement a logistic regression model that is suited for large scale genomic data and which provides a joint estimation of SNP effects. The R Package bigstatsr includes efficient functions to train penalized logistic regression on large datasets.

We present a comprehensive comparative study of the C+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests. Penalized logistic regression consistently achieves higher predictive performance than T-Trees while being as fast as the C+T method. Predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs which values depends on disease architecture. Penalized logistic regression consistently achieves superior performance than the C+T method and this improvement was more pronounced when there are few effects located in nearby genomic regions such as HLA. When analysing SNP data from a case-control study,  penalized logistic regression provides an AUC of $89\%$ while the standard C+T method provides a smaller $AUC$ of $82\%$.

Instead of computing polygenic risk scores based on GWAS univariate summary statistics, we recommend to use penalized logistic regression as implemented in the R package bigsnpr to obtain more accurate risk scores.

\\
\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
\textbf{Supplementary information:}
}

@@

In order to simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}. We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only chromosome 6), the size of the training set ($\{1000, 2000, \dots, 6000\}$), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace). We also use two different models to compute liability scores: a ``simple'' model with linear effects only, and a ``fancy'' model which combines linear, dominant and interaction-type effects.
For the ``simple'' model, we compute $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}}$$ where $w_j$ are weights generated with a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$ and $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs). 
For the ``fancy'' model, we separate the causal SNPs in three equal sets $S_\text{causal}^{(1)}$, $S_\text{causal}^{(2)}$ 
and $S_\text{causal}^{(3)}$ ($S_\text{causal}^{(3)}$ is further separated in two equal sets, $S_\text{causal}^{(3.1)}$ and $S_\text{causal}^{(3.2)}$). 
We then compute $$y_i = \underbrace{\sum_{j\in S_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in S_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|S_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction}$$ 
where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $S_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|S_\text{causal}^{(q)}\right|\right\}\right\}$. 
Note that for the interaction part of the model, we scale interactions, not the raw allele counts, so that corresponding SNPs still have a marginal effect. 
For both models, the values of $y_i$ are then standardized so that they have a variance equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$, where $\epsilon \sim N(0, 1 - h^2)$.

@@

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of a Genome-Wide Associations Studies (GWAS), referred to as summary statistics.
For the C+T method, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part). 
The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another, which form a set of SNPs denoted $S_\text{clumping}$. 
Then, thresholding (T) consists in removing SNPs with p-values larger than a threshold $t$ to be determined. 
Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$\rm{PRS}_i = \sum_{j \in S_\text{clumping}} \mathds{1}\{p_j < t\} \cdot \beta_j \cdot G_{i,j}$$ where $\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS. In this study, we report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb.

For the C+T method, we report three different scores of prediction: one including all the SNPs (remaining after clumping), one including only SNPs that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$), and one that maximizes the AUC for these two thresholds and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}). 
As we report the optimal threshold based on the test set, the corresponding AUC value is an upper bound of the AUC for the C+T method.
We call these three reported scores respectively ``PRS-all'', ``PRS-stringent'' and ``PRS-max'' in the results.

T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests \cite[]{breiman2001random} that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium in GWAS data \cite[]{botta2014exploiting}. We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}, except that we use 100 trees instead of 1000 because because using 1000 trees provides a minimal increase of AUC while requiring a disproportionately long processing time (e.g.\ AUC of 81.5\% instead of 81\%, data not shown). %We call this method ``T-Trees'' in the results.

Finally, for the penalized logistic regression, we solve: $$\argmin_{\beta_0, \beta}(\lambda, \alpha)\left\{\underbrace{\frac{1}{n}\sum_{i=1}^n \log\left(1+e^{-y_i (\beta_0+x_i^T\beta)}\right)}_\text{Loss function} + \underbrace{\lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right)}_\text{Penalization}\right\}$$ where, in this study, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict and $\lambda$ is a regularization parameter that needs to be determined. 
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse and therefore more interpretable models (corresponds to $\alpha = 1$); 
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. In this study, we use elastic-net regularization with $\alpha = 0.5$.%, without trying to tune (e.g.\ by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.

To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in the R package bigstatsr \cite[]{prive2017efficient}. 
This type of algorithm builds predictions for many values of $\lambda$ (typically a ``regularization path'' of 100 values). To obtain an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we called Cross-Model Selection and Averaging (CMSA). First, this procedure separates the training set in $K$ folds (e.g.\ 10 folds). 
Secondly, in turn, each fold is considered as an inner validation set and the other ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which minimizes the log loss on the inner validation set is determined and the vector of coefficients corresponding to the previous vector of scores is selected. Finally, the $K$ resulting probabilities for each individual are averaged. 
We also use this procedure to derive an early stopping criterion so that the algorithm does not need to evaluate the whole regularization paths, making this procedure much faster.
Because of the L1-regularization, the resulting vectors of coefficients are typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts. We refer to this method as ``logit-simple'' in the results section.

In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct a separate dataset with, for each SNP variable, two more variables coding for recessive and dominant effects. This results in a dataset with 3 times as many variables as the initial one, on which we can apply penalized logistic regression with the CMSA procedure, as described previously. We call this method ``logit-triple'' in the results.
